### 2010-04-13 thesis/SensorMLOnSenseTile/trunk/thesis/SensorMLonSenseTileThesis.tex
latest

Sensor systems are increasingly deployed to provide sensing information for a myriad of applications. These sensor systems contain many different sensor types and are developed by communities working in different domains. The heterogeneity of these systems means that interoperability is difficult and expensive. Reliably processing sensor observations retrieved from these heterogeneous systems is also problematic. A standard way to retrieve sensor data and a standard way to describe sensors are needed to address the current problem of sensor system heterogeneity.

Standards are being developed to address the above issues. Two such standards are the Sensor Model Language (SensorML)\cite{SMLref} specification and the Sensor Observation Service (SOS)\cite{SOSref} specification. SensorML is a proposed standard for sensor description and processing of sensor observations. SensorML specifies models and XML Schema for describing sensors, the process of sensor measurement and the post processing of sensor observations. The SOS specification is a proposed standard for retrieving sensor information and sensor observations using Web Services.

@@

Developing robust sensor systems is also a natural concern. There are now formal verification based software development processes \cite{Kiniryref} with the goal of developing very robust systems.  Currently there is no widely available formal definition of sensors or sensor data processing. Formal sensor descriptions allows sensor system development with such a verification based process and their development is examined also in this thesis.

@@

Buisness Object Notatation(BON)\cite{BONref} isi used to formalize SensorML sensor descriptions in this thesis. BON is a notation and a method for Object Oriented systems analysis and design. The BON to SensorML mapping described in thesis allows SensorML sensor descriptions to then be refined to Java as described in \cite{Kiniryref} as part of developing a robust Sensor System.

@@

\section{Sensor Web Enablement}\label{SWESec}
As described in the introduction the SensorML and SOS specifications are developed as part of the Open Geospatial Consortium (OGC) Sensor Web Enablement (SWE) initiative \footnote{http://www.opengeospatial.org/projects/groups/sensorweb}. This initiative is establishing interfaces and protocols to enable a standardized Sensor Web. It is hoped that this allows for easier integration of sensor applications across sensor networks. The OGC SWE provides a framework of open standards for use with Web-connected sensors and sensor systems. There are seven main specifications currently:

@@

SWE services based on the aboved specifications allow the discovery of sensors, access to the sensor data, as well as subscription to alerts, and tasking of sensors to control observations. For a good overview of SWE and it's use in Sensor Web Networks see \cite{SWEArchref}.  Along with with the SensorML and SOS specification, the O\&M specification is also used in developing the SenseTile Web Service as it defines the encoding of sensor observations.

@@

In SensorML sensors are modeled as processes that convert real phenomena to data. Processes take inputs, process them and result in one or several outputs. For example, a process can take a measurement generated by a thermistor which might be a resistance value and transform it to a celcius value. Processes can also be connected together in process chains to allow the sensor data to be processed by several processes sequentially. These processes and process chains are modeled in SensorML as shown in the  below SensorML Conceptual Model \ref{fig:SMLConceptualModel}.

@@

Processes in SensorML derive from AbstractProcess with it's inputs, output, parameters and metaDataGroup properties. The inputs and outputs represent data connections to other processes. Parameters represent inputs other than sensor measurements or outputs of other processes. They are used to configure the process or provide other inputs that are not sensor measurements. For example a parameter might be the latency time of a measurement needed for a calculation. Finally each process provides sensor metadata as modeled by the MetaDataGroup.

@@

The SensorML specification currently contains the Sensor Web Enablement (SWE) Common specification. This specification defines basic types and data encodings used by the SWE specifications. It describes both primitive data types as well as aggregate types such as arrays and records. There are a large number of types defined in this specification, too many to show here. Below is a Conceptual Model of some of the simple data types.

@@

The data types in \ref{fig:SWESimpleConceptualModel} are scalar primitive values. They are the basis for the inputs, outputs and parameters in SensorML processes. Quantity models a floating point number. Count models a counting value or integer.  Boolean models a property that can be true or false. These types contain a number of properties that further define the type. The uom property provides a unit of measurement that indicate how the value should be interpreted.  A constraint property allows a value range or a enumerated list of values to be defined for the type. The quality property allows for a measure of its quality, for example the confidence level of a value being correct expressed as a percentage. Aggregation of these primitive data types is also provided by several types including DataRecord and DataArray.

@@

\section{Sensor Observation Service}\label{SOSSec}
SensorML is envisioned to be used in conjunction with a Sensor Observation Service as described in the SWE Architecture \cite{SOSref}. The SOS defines a web service interface for the discovery of sensors systems and the retrieval of observations from sensor systems. It organizes related sensor observations into a collection called an observation offering.

The SOS specification uses four main entities to describe the SOS's operation. The entities are Data Consumers, Data Producers, OGC Catalogs and the SOS itself. Data Consumers are clients of the SOS that request sensor observations and sensor metadata. Data Providers are entities have to access to the sensor and have enough processing power to generate an XML encoding of the sensor's observations. These observations are sent over a network to the SOS. The OGC Catalog is used to find SOS instances. SOS instances act as access points to a Sensor Web Network and provide the Web Service for retrieval of sensor observations. The figure \ref{fig:SOSoperationContext} below shows the entities described in the SOS in an operation context.
\begin{figure}[h]
\framebox{\scalebox{0.65}{\includegraphics{SOSoperationContext.png}}}
\caption{SOS Operational Context}\label{fig:SOSoperationContext}
\end{figure}

@@

These operations are encoded in XML and are intended to be carried http  operations. The sensor observations contained in  GetObservation and InsertObservation operations are encoded according to the OGC O\&M Observations \cite{OMref}. Sensor systems metadata contained in DescribeSensor and RegisterSensor is defined in SensorML or TML.

@@

The OGC Catalog Service is not explored in this thesis. The SOS Operation Context diagram shows a discovery process involving the  Data Consumer and a OGC Catalog Service (CS-W) catalog. The Data Consumer uses a CS-W catalog to find SOS instances. The GetCapabilites are then used toward each of the SOS instances returned to find sensor systems that are of interest to the Data Consumer. The CS-W catalog is covered in a separate specification \cite{OGCcatref}. This not covered in any real detail in the SOS specification.

@@

Thermistor?

\section{UCD Sensor Board Driver}\label{SensorBoardDriverSec}

\section{VAST SensorML Engine}\label{VastSensorMLEngineSec}
The VAST Team at the University of Alabama in Huntsville (UAH)  developed an open source Java based SensorML Processing Engine. This software provides types that implement SensorML processes and process  chains. The SensorML Processing Engine parses SensorML models and instantiates the required objects to performed the intended processing. The code is found at \footnote{http://code.google.com/p/sensorml-data-processing} and is referred to as the VastSMLEngine in rest of the thesis. This code is dependent on an implementation of the SWE Common data types described in the section on SensorML \ref{SMLsection}. This code is found at \footnote{http://code.google.com/p/swe-common-data-framework}.

@@

The SenseTile Web Service is based on the Sensor Observation Service (SOS) as described previously. The two main entities in the SOS Specification, the Sensor Data Provider and the SOS,  are used to structure the SenseTile Web Service. These entities are implemented as seperate software components that together provide the service. The Sensor Data Provider component will be referred to as the DataProvider in this design.

The two components are run as seperate processes on the SenseTile processor unit to allow a more flexible network architecture.  This network architecture is described in more detail in section \ref{subsec:SenseTile WebService Network Architecture}. The basic system structure is shown in figure \ref {fig:Deployment_sensetile} .

@@

\newpage
\subsection{Supported Scenarios}
This section provides UML sequence diagrams to show the how the DataProducer and SOS interact to provide the sensor to the DataConsumer client.

\subsubsection{Generate Observations from Sensor Data}
The DataProducer accesses data from the sensor board 
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{InsertObs.png}}}
\caption{Generate Observation}\label{fig:InsertObs}
\end{figure}
 \begin{enumerate}
\item -
\item -
\item -
\end{enumerate}

\subsubsection{Get Sensor Description from SOS}
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{GetMeta.png}}}
\caption{Find SOS sensor}\label{fig:GetMeta}
\end{figure}
 \begin{enumerate}
\item -
\item -
\end{enumerate}
\subsubsection{Get  Sensor Observation from SOS}
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{GetObs.png}}}
\caption{Get Observation from Sensor}\label{fig:GetObs}
\end{figure}
 \begin{enumerate}
\item -
\end{enumerate}

@@

@@

\chapter{ Conclusions and Future Work}
This thesis analyses the role of SensorML and SOS specifications in Sensor Systems. The UCD CASL SenseTile System is used as a case study for the use of these specifications. A SensorML model of SenseTile and a SenseTile Web Service prototype.

SensorML address the issues of interoperabilty of Sensor Systems
rich description of a sensor systems. Hard to think of any system that could no be described by SensorML. Weather systems to Geloloaction services. 
Accurate elements. Processing data. Metadata.
The supporting SWE Common data types are many data types.
SensorML Frameworks.
Precisely define its data.
It also is bases for developement for Sensor Ontology work.\footnote{http://www.calebgoodwin.com/SEMANTIC.html}
SensorML is relatively intuitive but it is difficult to know if one has developed a "good" SensorML model. There are examples and tutorials to help.

@@

@@

\newpage
\begin{thebibliography}{99}
\bibitem{OSWAref} Chu, Kobialka,  Durnota, and  Buyya, Open Sensor Web Architecture: Core Services
\bibitem{SMLref}Open Geospatial Consortium Inc., OpenGIS Sensor Model Language (SensorML) Implementation Specification, 2007
\bibitem{SOSref}Open Geospatial Consortium Inc.,  Sensor Observation Service, 2007
\bibitem{Kiniryref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{OMref}Open Geospatial Consortium Inc., Observations and Measurements – Part 1 - Observation schema, 2007
\bibitem{SWEArchref}Open Geospatial Consortium Inc., OGC Sensor Web Enablement Architecture, 2008
\bibitem{SMLTutorialref}Open Geospatial Consortium Inc., Using SensorML to describe a
Complete Weather Station , 2006
\bibitem{ProcessTutorialref}Open Geospatial Consortium Inc.,
\bibitem{OGCcatref}Open Geospatial Consortium Inc., OpenGIS® Catalogue Services Specification, 2007
\bibitem{BONref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{Meyerref}Bertrand Meyer., Object-Oriented Software Construction, 1997

>>>>

Sensor systems are increasingly deployed to provide sensing information for a myriad of applications. These sensor systems contain many different sensor types and are developed by communities working in different domains. The heterogeneity of these systems means that interoperability is difficult and expensive. Reliably processing sensor observations retrieved from these heterogeneous systems is also problematic. Standard ways to describe sensors and to retrieve sensor data and are needed to address the current problem of sensor system heterogeneity.

Standards are being developed to address the above issues. Two such standards are the Sensor Model Language (SensorML) specification\cite{SMLref} and the Sensor Observation Service (SOS) specification\cite{SOSref}. SensorML is a proposed standard for sensor description and processing of sensor observations. SensorML specifies models and XML Schema for describing sensors, the process of sensor measurement and the post processing of sensor observations. The SOS specification is a proposed standard for retrieving sensor information and sensor observations using Web Services.

@@

@@

Currently there is no widely available formal definition of sensors or sensor data processing. There are now formal verification based software development processes \cite{Kiniryref} with the goal of developing very robust systems.  Formal sensor descriptions allows sensor system development with such a verification based process. Business Object Notatation(BON)\cite{BONref} is used in \cite{Kiniryref} to formally specify the system being developed. BON is a notation and a method for analysis and design of Object Oriented systems.

This thesis looks at how BON can be used to formally specify SensorML. The BON to SensorML mapping described in thesis allows SensorML sensor descriptions to then be refined to Java as described in \cite{Kiniryref} as part of developing a robust Sensor System.

@@

\section{Sensor Web Enablement}\label{SWESec}
As described in the introduction the SensorML and SOS specifications are developed as part of the Sensor Web Enablement (SWE) initiative \footnote{http://www.opengeospatial.org/projects/groups/sensorweb}. This Open Geospatial Consortium (OGC) initiative is establishing interfaces and protocols to enable a standardized Sensor Web. This allows for easier integration of sensor applications across sensor systems. SWE provides a framework of open standards for use with Web-connected sensors and sensor systems. There are seven main specifications currently:

@@

SWE services based on the above specifications allow the discovery of sensors, access to the sensor data, as well as subscription to alerts, and tasking of sensors to control observations. For a good overview of SWE and it's use in Sensor Web Networks see \cite{SWEArchref}.  Along with with the SensorML and SOS specification, the O\&M specification is also used in developing the SenseTile Web Service as it defines the encoding of sensor observations.

@@

\subsection{SensorML Core Model}
In SensorML sensors are modeled as processes that convert real phenomena to data. Processes take inputs, process them and result in one or several outputs. For example, a process can take a measurement generated by a thermistor which might be a resistance value and transform it to a celcius value. Processes can also be connected together in process chains to allow the sensor data to be processed by several processes sequentially. These processes and process chains are modeled in SensorML as shown in the  below SensorML Conceptual Model \ref{fig:SMLConceptualModel}.

@@

Processes in SensorML derive from AbstractProcess with it's inputs, output, parameters and metaDataGroup properties. The inputs and outputs represent data connections to other processes. Parameters are used to configure the process or provide other inputs that are not sensor measurements. For example a parameter might be the latency time of a measurement needed for a calculation.

Finally each process provides sensor metadata as modeled by the MetaDataGroup.

@@

\subsection{Common Data}
The SensorML specification currently contains the Sensor Web Enablement (SWE) Common specification. This specification defines basic types and data encodings used by the SWE specifications. It describes both primitive data types as well as aggregate types such as arrays and records. There are a large number of types defined in this specification, too many to show here. Below is a Conceptual Model of some of the simple data types.

@@

The data types in \ref{fig:SWESimpleConceptualModel} are scalar primitive values. They are the basis for the inputs, outputs and parameters in SensorML processes. Quantity models a floating point number. Count models a counting value or integer. Boolean models a property that can be true or false.

These types contain a number of properties that further define the type. The uom property provides a unit of measurement that indicate how the value should be interpreted.  A constraint property allows a value range or a enumerated list of allowed values to be defined for the type. The quality property allows for a measure of its quality, for example the confidence level of a value being correct expressed as a percentage. Aggregation of these data types is also provided by several types including DataRecord and DataArray.

@@

\subsection{VAST SensorML Engine}\label{VastSensorMLEngineSec}
The VAST Team at the University of Alabama in Huntsville (UAH) has developed an open source Java based SensorML Processing Engine. This software provides types that implement SensorML processes and process  chains. The SensorML Processing Engine parses SensorML models and instantiates the required objects to performed the intended processing. The code is found at \footnote{http://code.google.com/p/sensorml-data-processing} and is referred to as the VastSMLEngine in rest of the thesis. This code is dependent on an implementation of the SWE Common data types described in the section on SensorML \ref{SMLsection}. This code is found at \footnote{http://code.google.com/p/swe-common-data-framework}.

\section{Sensor Observation Service}\label{SOSSec}
The SOS specifications defines a web service interface for the retrieval of sensor descriptions and the sensor observations from sensor systems. It organizes related sensor observations into a collection called an observation offering.   
SensorML is used in conjunction with a Sensor Observation Service as described in the SWE Architecture \cite{SOSref} for sensor descriptions.

The SOS specification uses four main entities to describe the SOS's operation. The entities are Data Consumers, Data Producers, OGC Catalogs and the SOS itself. Data Consumers are clients of the SOS that request sensor observations and sensor metadata. Data Providers are entities have to access to the sensor and have enough processing power to generate an XML encoding of the sensor's observations. These observations are sent over a network to the SOS. The OGC Catalog is used to find SOS instances. SOS instances act as access points to a Sensor Web Network and provide the Web Service for retrieval of sensor observations. The figure \ref{fig:SOSoperationContext} below shows the entities described in the SOS in an operational context.
\begin{figure}[h]
\framebox{\scalebox{0.65}{\includegraphics{SOSoperationContext.png}}}
\caption{SOS Operational Context}\label{fig:SOSoperationContext}
\end{figure}

@@

These operations are encoded in XML and are intended to be carried http  operations. The sensor observations contained in  GetObservation and InsertObservation operations are encoded according to the OGC O\&M Observations \cite{OMref}. Sensor systems metadata contained in DescribeSensor and RegisterSensor is defined in SensorML.

@@

The OGC Catalog Service shown in the SOS Operation Context diagram \ref{fig:SOSoperationContext} is not explored in this thesis. The Data Consumer uses a CS-W catalog to find SOS instances. The GetCapabilites operation is then used toward each of the SOS instances found to find sensor offerings that are of interest to the Data Consumer. The CS-W catalog is covered in a separate specification \cite{OGCcatref} and is not covered in any real detail in the SOS specification.

@@

The TMP175 and TMP75 are digital temperature sensors
that are optimal for thermal management and thermal
protection applications. The TMP175 and TMP75 are
Two-Wire and SMBus interface-compatible, and are
specified over a temperature range of −40°C to +125°C.

\subsection{UCD Sensor Board Driver}\label{SensorBoardDriverSec}

@@

The SenseTile Web Service is based on the Sensor Observation Service (SOS) as described previously. The two main entities in the SOS Specification, the Sensor Data Provider and the SOS,  are used to structure the SenseTile Web Service. These entities are implemented as separate software components that together provide the service. The Sensor Data Provider component will be referred to as the DataProvider in this design.

The two components are run as separate processes on the SenseTile processor unit to allow a more flexible network architecture.  This network architecture is described in more detail in section \ref{subsec:SenseTile WebService Network Architecture}. The basic system structure is shown in figure \ref {fig:Deployment_sensetile} .

@@

@@

\newpage
\subsection{SenseTile Web Service Scenarios}
This section contains UML sequence diagrams to show the how the DataProducer and SOS interact to provide the sensor to the DataConsumer client. There are three main scenarios supported by the SenseTile Web Service

\subsubsection{Generate Observations from Sensor Data}
The DataProducer accesses data from the sensor board and updates the SOS with sensor observations.
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{InsertObs.png}}}
\caption{Generate Observation}\label{fig:InsertObs}
\end{figure}
 \begin{enumerate}
\item At start up the DataProducer registers the SensorBoard System and the Converter System with the SOS in separate RegistorSensor operations. The SOS creates sensor offerings for each of the Systems and stores the SensorML descriptions of the Systems provided in the RegistorSensor operation. It then returns a unique id (URN) in the RegistorSensor response for use by the DataProducer for subsequent communication with the SOS.
\item The DataProducer reads Sensor Board data packets from the SensorBoardDriver.
\item The DataProducer updates the SOS with sensor observations with the InsertObservation operation.
\end{enumerate}

\subsubsection{Get Sensor Description from SOS}
The SOS allows DataConsumers to obtain SensorML descriptions of the SesneTile Systems.
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{GetMeta.png}}}
\caption{Get Sensor Descriptions}\label{fig:GetMeta}
\end{figure}
 \begin{enumerate}
\item The DataConsumer requests information on sensor offerings provided by the SOS using the GetCapabilities operation.
\item Using the information in the GetCapabilites response the DataConsumer requests a description of a sensor system that is providing an offering using the DescribeSensor operation. The SOS returns the SensorML description of the sensor system to the DataConsumer.
\end{enumerate}

\subsubsection{Get  Sensor Observation from SOS}
\begin{figure}[h]
\centering
\framebox{\scalebox{0.65}{\includegraphics{GetObs.png}}}
\caption{Get Observation from Sensor}\label{fig:GetObs}
\end{figure}
 \begin{enumerate}
\item Using a sensor identity received in a GetCapability response the DataConsumer requests a sensor observation. The SOS returns the latest O\&M observation to the DataConsumer.
\end{enumerate}

@@

\chapter{ Conclusions and Future Work}
This thesis analyses the role of the SensorML and the SOS specifications in Sensor Systems. The UCD CASL SenseTile System is used as a case study for the use of these specifications. A SensorML model of the SenseTile system is developed to evaluate SensorML. A SenseTile Web Service prototype is used to evaluate the SOS specification which uses SensorML to provide sensor descriptions to clients.

SensorML provides a good model of sensor systems in terms of a an intuitive process concept.  to describe of a sensor systems. It is hard to think of any system that could no be described by SensorML due to number of data types that can. Weather systems to Geloloaction services. SensorML address the issues of interoperabilty of Sensor Systems

For descriptions there good Metadata.

SensorML provides good support for the reliable processing of sensor data. The data types have several properties that can reference definitions to precisely describe their meaning.  The accuracy of the sensor data can be modelled and provided used by systems to process the data with these variables include to get a better result.

Processing data. 
SensorML Frameworks. DataProducer.

It also is bases for development for Sensor Ontology work.\footnote{http://www.calebgoodwin.com/SEMANTIC.html}
SensorML is relatively intuitive but it is difficult to know if one has developed a "good" SensorML model. There are examples and tutorials to help.

@@

SOS Scalable hierarcy of instances.

@@

\newpage
\begin{thebibliography}{99}
\bibitem{OSWAref} Chu, Kobialka,  Durnota, and  Buyya, Open Sensor Web Architecture: Core Services
\bibitem{SMLref}Open Geospatial Consortium Inc., OpenGIS Sensor Model Language (SensorML) Implementation Specification, 2007
\bibitem{SOSref}Open Geospatial Consortium Inc.,  Sensor Observation Service, 2007
\bibitem{Kiniryref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{OMref}Open Geospatial Consortium Inc., Observations and Measurements – Part 1 - Observation schema, 2007
\bibitem{SWEArchref}Open Geospatial Consortium Inc., OGC Sensor Web Enablement Architecture, 2008
\bibitem{SMLTutorialref}Open Geospatial Consortium Inc., Using SensorML to describe a
Complete Weather Station , 2006
\bibitem{ProcessTutorialref}Open Geospatial Consortium Inc.,
\bibitem{OGCcatref}Open Geospatial Consortium Inc., OpenGIS® Catalogue Services Specification, 2007
\bibitem{BONref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{Meyerref}Bertrand Meyer., "Object-Oriented Software Construction", 1997
